Co-evolution of lexical meaning \& pragmatic use

### 2018-03-09 paper/cogsci-submission2.tex
restructured sections

The replicator mutator dynamic is an abstract description of population-level changes. It
abstracts away from specific assumptions about how agents interact. To understand this better,
the following looks at each component in isolation: first replication, then mutation.

@@

In the following application to language evolution, we prefer a cultural interpretation of the
mutator component as the perturbation induced by how easy different linguistic traits are
learnable from observation of language use. Nothing much hinges on whether the replicator
component is interpreted as a biological process or a cultural process. Fitness-based selection
may either be thought of as the outcome of actual reproductive success (increased number of
offspring which learn a language from their single parent), or as a process of cultural
evolution, such as conditional imitation for instance. We believe that this flexibility of
interpretation is an asset. It abstracts away from specific assumptions and focuses on a
general, albeit abstract description of fitness-based selection in general. Still, to have
something concrete in mind when interpreting the following application, it is also important to
be concrete. Appendix~\ref{sec:deriv-repl-mutat} therefore derives the replicator mutator
dynamic from a specific scheme of agent-level conditional imitation.

@@

Moving beyond an abstract example, our goal is to apply the RMD to investigate the evolution
of lexical representations and pragmatic behavior. To do so, we need to fix three things: (i) what the
relevant types are that populate $x$, (ii) how fitness derives from communicative success and (iii) how the
mutation matrix $Q$ is computed. These issues are addressed, one by one, in the following.

\subsubsection{Types: Lexica and pragmatic strategies}
\label{sec:languages+use}
Types are what evolution operates on. They define an agent's fitness, usually through a payoff
accrued in single interactions with other agents. Often types can be identified as the possible
acts in a game, e.g., either cooperating or defecting in a prisoner's dilemma. In other cases,
they may be thought of as general properties of an agent that influence her fitness, such as
being positive or negative in our previous example (whatever that means). For our present
purposes, types are identified by specific assumptions about their cognitive
make-up. More concretely, since we are interested in the evolutionary competition between different lexical representations and ways of deploying them in communication, a type is here defined as a pair consisting of a lexicon and a linguistic strategy of language use.

Agents play signaling games, in which the speaker wants to communicate a world state $s$ with a
message $m$ to a hearer who receives $m$ but does not know $s$
\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a
(possibly fuzzy) set of states. A linguistic behavior specifies a probabilistic speaker
rule (a probabilistic choice of message for each state) and a probabilistic hearer rule (a
probabilistic choice of state for each message) given a lexicon. There are many ways of making these general notions concrete. There are models of rational language use coming from a game-theoretic tradition \citep[e.g.,][]{benz:2006,benz+vRooij:2007,jaeger:2007b,franke+jaeger:2014} and ones from a Bayesian tradition \citep[e.g.,][]{frank+goodman:2012,goodman+stuhlmueller:2013,GoodmanFrank2016:Pragmatic-Langu}. We will not focus on the particular design choices that make up each approach (on this see \citealt{qing+franke:2015}). Instead, we will stress how the approach, common to all of these models, of pragmatic language use as {\em rational language use} can be embedded within the replicator mutator dynamic to answer evolutionary questions at the semantics-pragmatics interface. Here is what we will assume in the remainder of this paper.\footnote{Readers more familiar with the Bayesian tradition than the game-theoretic one should note that the definitions below follow the latter in assuming rational sender choice to consist in sending messages that maximize expected utility, rather than message informativity (see \citealt{qing+franke:2015} for discussion).}

@@

In words, an agent's linguistic behavior -- what message to send when or which message to interpret how -- is defined by (i) her reasoning level and (ii) her lexicon. In what follows, this is what we identify an agent's type with, in order to analyze under which conditions combinations of (i) and (ii) emerge through cultural evolution. Intuitively, literal behavior (level-$0$ reasoning) results from unreflected language use. Such agents produce and comprehend solely based on the truth-conditions of their lexicon. In a Gricean spirit, pragmatic behavior (level-$n+1$ reasoning) results from mutual reasoning about rational language use. Such agents produce and comprehend by reasoning about their interlocutor's likely interpretation/production patterns. Higher order behavior is therefore, even if in tendency rational, quite simple. It does not assume agents to know each other's types when interacting (e.g., what lexicon the interlocutor is using); they simply act based on a conjecture of their own behavior if they were in the role of their interlocutor.

@@

Under the replicator dynamic the proportion of type $i$ in a population will increase or
decrease as a function of its relative fitness $f_i$. In the context of language evolution,
fitness is usually associated with the ability to successfully communicate
with other language users from the same population
\citep[e.g.,][]{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Under a biological
interpretation the assumption is that organisms have a higher chance of survival and
reproduction if they are able to share and receive useful information via communication with
peers. Under a cultural interpretation the picture is that agents themselves strive towards
communicative success and therefore occasionally adapt or revise their behavior to achieve
higher communicative success (see \citealt[\S3.3]{benz+etal:2005b} for discussion).

The replicator equation gives us the means to make the ensuing dynamic precise, without
necessarily committing to a biological or cultural interpretation. As above, the proportion of
types in a given population is codified in a vector $\vec{x}$, where $x_i$ is the proportion of
type $i$. The fitness of type $i$ is its average expected communicative success, or
\emph{expected utility} (EU), given the frequencies of types in the current population:
\begin{align*}
  f_i = \sum_j x_j \text{EU}(t_i,t_j)\,.
\end{align*}
In signaling games, the expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
average success of $i$ when talking or listening to $j$. If, as standardly assumed, agents are speakers half of the
time, this yields:
\begin{align*}
  \text{EU}(t_i,t_j) = \nicefrac{1}{2} \, \text{EU}_S(t_i,t_j) + \nicefrac{1}{2} \, \text{EU}_H(t_i,t_j)\,,
\end{align*}
where $\text{EU}_S(t_i,t_j)$ and $\text{EU}_H(t_i,t_j)$ are the expected utilities for $i$ as a
speaker and as a hearer when communicating with $j$, defined as follows, where $n_i$ and $n_j$
are type $i$'s and type $j$'s pragmatic reasoning types and $L_i$ and $L_j$ are their lexica:
\begin{flalign*}
  & \text{EU}_S(t_i,t_j)  = \sum_s P(s)\sum_m S_{n_i}(m \mid s;L_i) \sum_{s'} H_{n_j}(s' \mid m;L_j)
  \delta(s,s') \,, \\
 & \text{EU}_H(t_i,t_j)  = \text{EU}_S(t_j,t_i)\,.
\end{flalign*}
As usual, we assume that agents are cooperative: $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.

@@

\subsubsection{Learnability}
\label{sec:learnability}
Languages are shaped not only by functionalist forces towards greater communicative success. Another
important factor is the fidelity with which linguistic knowledge is transmitted. Among others,
linguistic production can be prone to errors, states or messages may be perceived incorrectly,
and multiple languages may be compatible with the data learners are exposed to. These sources
of uncertainty introduce variation in the transmission of linguistic knowledge from one
generation to the next. In particular, learning biases in the iterated transmission process can
influence language evolution substantially.

In biological evolution, where types are expressed genetically, transmission infidelity comes
into the picture through infrequent and mostly random mutation and genetic drift \citep{kimura:1983}. However, an agent's
lexicon and pragmatic reasoning behavior is likely not inherited genetically. They need to be
learned from observation.  Concretely, when agents learn from type $j$, they observe the overt linguistic behavior of type $j$ and need to infer the covert type that most likely produced the observed behavior. As in biology, this transmission process is not perfectly accurate.

@@

Following \citet{griffiths+kalish:2005,griffiths+kalish:2007} we model language acquisition as a process of Bayesian
inference in which learners combine the likelihood of a type producing the witnessed learning
input with prior inductive biases. In a Bayesian setting these biases can be codified in a prior
$P \in \Delta(T)$, which reflects the amount of data a learner requires to faithfully acquire
the language of the teacher \citep[cf.][450]{griffiths+kalish:2007}. Put differently, types favored by the prior are easier to learn because they require less data than less favored types to be inferred, even if they are both compatible with the learning data at hand. The extent of the prior's
influence has been shown to heavily depend on the learning strategy assumed to underlie the
inference process. On the one hand, early simulation results suggested that weak biases could
be magnified by exposing learners to only small data samples
\citep[e.g. in][]{brighton:2002}. On the other, \citeposs{griffiths+kalish:2007} mathematical
characterization showed that iterated learning converges to the prior in the limit, i.e., that
the resulting distribution over types corresponds to the learners' prior distribution and
is not influenced by the amount of input given to them. This difference in predictions can be
traced back to differences in the selection of hypotheses from the posterior. Griffith \&
Kalish's convergence to the prior holds for learners that sample from the posterior. More
deterministic strategies such as the adoption of the type with the highest posterior
probability, so-called {\it maximum a posterior estimation} (MAP), increase the influence of
both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we
use a parameter $l\ge1$ to modulate between posterior sampling and the MAP strategy. When
$l = 1$ learners sample from the posterior. The learners' propensity to maximize the posterior
grows as $l$ increases.

The way the posterior over types is formed and how learners employ it are important design choices. There are other alternatives as well. For instance, instead of adopting a single type, learners could keep their uncertainty over types and base their production behavior on the posterior over types they form \citep{burkett+griffiths:2010}. A straightforward operationalization of this idea would then see production defined as a two-step process wherein a speaker first samples from her posterior and then produces utterances based on the type she sampled, as done in \citealt{kirby+etal:2015}. Beyond adding complexity to the model, a disadvantage of this approach is that it could happen that a speaker that first uttered a Hungarian utterance speaks Quechua in the next, so to speak. We stick to the simpler assumption of a single type being acquired, which we now define.

@@

\subsection{Example}\label{subsec:example}
The case study in Section~\ref{sec:si-case-study} shows how these components and their interaction can shed light on the evolution of scalar implicatures. Before moving on, this section illustrates the model's workings in a constrained type space to better trace the effects that pressure for communicative success and learnability can have on an evolving population. To this end, we focus only on the four types from Section~\ref{sec:languages+use}. Recall that these types are made up of a combination of either lexicon $\Llack$ or $\Lbound$ and either literal (level-$0$) or pragmatic (level-$1$) behavior.

  % \path[->, thick, densely dashdotted, Steelblue] (0.75,0.75) edge [bend left] node[sloped,below] {\footnotesize learning\textsubscript{flat}} (2.25,2.45);
  % \path[->, thick, densely dashdotted, Steelblue] (4.25,0.75) edge [bend right] node[sloped,below] {\footnotesize leaning\textsubscript{flat}} (2.75,2.45);
  % \path[->, thick, densely dashdotted, Steelblue] (0.75,4.25) edge [bend right] node[sloped,above] {\footnotesize leaning\textsubscript{flat}} (2.25,2.75);
  % \path[->, thick, densely dashdotted, Steelblue] (4.25,4.25) edge [bend left] node[sloped,above] {\footnotesize leaning\textsubscript{flat}} (2.75,2.75);

  %   \draw[->, thick, goldish] (0.75,-0.25) -- (2.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
  %   \draw[<-, thick, goldish] (2.75,-0.25) -- (4.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
  %   \draw[->, thick, goldish] (-0.25,0.75) -- (-0.25,4.25) node[midway, above,rotate=90] {\footnotesize learning\textsubscript{biased}};
  %   \draw[->, thick, goldish] (5.25,0.75) -- (5.25,4.25) node[midway, below,rotate=90] {\footnotesize learning\textsubscript{bias}};
  %   \draw[->, thick, goldish] (0.75,5.25) -- (4.25,5.25) node[midway, above] {\footnotesize learning\textsubscript{biased}};

  %   \end{tikzpicture}
\caption{Sketch of dynamics on a two-dimensional type space with four types. The $x$-axis represents the proportion of types with $\Llack$; the $y$-axis that of pragmatic ones. Arrows sketch population trajectories under functional pressure (replicator steps) and learnability pressure (mutator steps) with either a flat prior or one that favors $\Llack$.}
\label{fig:sketch}
\end{figure}

A population will be made up of frequencies of these four types. We may think of such a population as a point in a two-dimensional space. The grey dotted square in Fig.~\ref{fig:sketch} outlines this space. The space's first dimension determines the proportion of types that lexicalize ambiguous $\msome$ (the $x$-axis in Fig.~\ref{fig:sketch}). That is, the proportion of types with $\Llack$. The second dimension determines the proportion of pragmatic language users (the $y$-axis in Fig.~\ref{fig:sketch}). Populations at the corners of this space are monomorphic; all agents in them are of a single type. If we move from away from edges toward the center of the space, then this two-dimensional representation underspecifies the exact proportions of types found in a population. The space's center, for instance, could be composed 50\% of pragmatic $\Llack$ and of an equal proportion of literal $\Lbound$. But it could equally well be a population that has each type with the same frequency. This level of granularity will nevertheless be sufficient for our purposes, as we are only interested in sketching rough trajectories after replicator or mutator steps.

From Section~\ref{sec:languages+use}, we already know some things about this type space. When it comes to communicative success: types with $\Lbound$ have a functional advantage over $\Llack$; literal $\Lbound$ has a slight advantage over pragmatic $\Lbound$; all have a sizable advantage over literal $\Llack$. The red arrows in Fig.~\ref{fig:sketch} sketch out these relationships. They depict the rough trajectories we can expect if there is only pressure for communicative success. How much each replicator step moves a population through this space ultimately depends on the rationality parameter $\lambda$. In particular, note that functional differences decrease as $\lambda$ increases: if $\lambda$ is sufficiently high, all types except for literal $\Llack$ speakers will show a strong tendency to associate $\ssome$ with $\msome$ and $\sall$ with $\mall$. As a consequence, when it comes to learnability and transmission fidelity, the likelihood that any of these three types produces a given datum $d$ will be approximately equal. In other words, if $\lambda$ is high the likelihood offers little differentiation between pragmatic $\Llack$ and any disposition to use $\Lbound$ for learners. We can therefore expect that, if the prior is flat, no type is passed on very faithfully in this space. This would lead populations to gravitate toward the space's center after mutator steps. This is sketched by the blue dash-dotted arrows in Fig.~\ref{fig:sketch}. Neither prior nor likelihood alone, but their combination, are what determines the effects of (iterated) Bayesian learning. If, instead of a flat prior, there is a prior that favors, say, $\Llack$ over $\Lbound$ we can expect mutator steps to be skewed toward the upper-half of the space. This is suggested by the purple arrows in Fig.~\ref{fig:sketch}. Taking stock, the effects of iterated learning will depend not only on $\lambda$, affecting the likelihood, but also on the prior (how strong of a bias there is) and the degree to which learners tend to maximize the posterior (controlled by parameter $l$).

\begin{figure}[t]
\centering
\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig-contourb105k5.png}
\caption{Dynamics on a two-dimensional type space with four types. The $x$-axis represents the proportion of types with $\Llack$; the $y$-axis that of pragmatic ones. Arrows indicate directionality of trajectories after (a) replicator steps and (b) mutator steps with a prior that favors types with $\Llack$ by a factor of $1.05$ over those with $\Lbound$. Colored contours show the proportion of pragmatic $\Llack$ users after a step \tb{I don't think the contours are adding much, so they could be removed. Then again, I think they are rather non-intrusive, look nice, and illustrate how the step-size, depending on where we are in the space, varies.}}

Fig.~\ref{fig:quiver} shows actual evolutionary trajectories in this type space. The replicator steps make the functional disadvantage of $\Llack$ against $\Lbound$ apparent, particularly for its literal users and/or low $\lambda$. They also highlight that the functional difference between literal and pragmatic $\Lbound$ is small. Under functional pressure only, populations move toward the lower part of the square but the steps become small as the lower edge, representing populations composed only  of $\Llack$ users, are reached. As for the mutator diagrams, there are two main things to note. As expected, a small bias for a type will lead to larger proportions to adopt it, provided that likelihood offers little to no differentiation between types. This is the situation the right-most space in Fig.~\ref{fig:quiver} depicts. However, the bias is far from absolute. If, instead, certain types are more likely to be recovered based on the linguistic input they produce, they will be passed on more faithfully. This is why, for instance, populations move toward a majority of $\Llack$-types for $\lambda = 5$ and $l = 10$.

Fig.~\ref{fig:sketch} and \ref{fig:quiver} depict replicator and mutator steps in isolation to better showcase their effects. Their combination will yield different outcomes, which we now turn to in a case study on scalar implicatures. \tb{Also: In this type space RMD way too similar to MD to make it worth the ink. It raises more questions than it answers so I left it out as the plots look identical to those in Fig.~3b}

@@

In the following we consider a specific application of the model from Section~\ref{sec:model}
which allows us to address the question if or when scalar inferences might (not) lexicalize. We
consider what is perhaps one of the simplest non-trivial setups that speak to this matter and
reflect on its limitations in Section~\ref{sec:discussion}. The setup is introduced in
Section~\ref{sec:setup:-stat-conc}. Section~\ref{sec:simulation-results} describes simulations
and their results.

% Our initial question about the division of labor between semantics and pragmatics can be narrowed to the case of scalar implicatures by asking for a justification for the lack of lexical upper-bounds in weak scalar alternatives. That is, we ask why lexical meanings that lack upper-bounds and convey it pragmatically are regularly selected for over alternatives such as that of codifying the bound semantically. More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} and {\em big} were truth-conditionally incompatible with stronger alternatives such as {\em hot}, {\em and}, {\em all} and {\em huge} to avoid misunderstanding?  This question is particularly striking considering the number of expressions that license such inferences across natural languages.

>>>>

The RMD is an abstract description of population-level changes. It abstracts away from specific
assumptions about how agents interact. To understand this better, the following looks at each
component in isolation: first replication, then mutation.

@@

In the following application to language evolution, we prefer a cultural interpretation of the
mutator component as the perturbation induced by how easy different linguistic traits are
learnable from observation of language use. Nothing of current relevance hinges on whether the
replicator component is interpreted as a biological process or a cultural process. We believe
that this flexibility of interpretation is an asset. It abstracts away from specific
assumptions and focuses on a general, albeit abstract description of fitness-based selection in
general. Still, to have something concrete in mind when interpreting the following application,
it is also important to be concrete. Appendix~\ref{sec:deriv-repl-mutat} therefore derives the
RMD from a specific scheme of agent-level conditional imitation.

@@

Our goal is to apply the RMD to the evolution of lexical representations and pragmatic
behavior. This requires determining three things: (i) what the relevant types are, (ii) how
fitness of a type derives from its communicative success and (iii) how the mutation matrix $Q$
is computed. These issues are addressed, one by one, in the following.

\subsubsection{Types: Lexica and pragmatic strategies}
\label{sec:languages+use}
Types are what evolution operates on. They define an agent's fitness, usually through a payoff
accrued in single interactions with other agents. Often types can be identified as the possible
acts in a game, e.g., either cooperating or defecting in a prisoner's dilemma. In other cases,
they may be thought of as general properties of an agent that influence her fitness, such as
being positive or negative in our previous example (whatever that means). Since we are
interested in the evolutionary competition between different lexical representations and ways
of deploying them in communication, a type is here defined as a pair consisting of a lexicon
and a linguistic strategy of language use.

Agents play signaling games, in which the speaker wants to communicate a world state $s$ with a
message $m$ to a hearer who receives $m$ but does not know $s$
\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a
(possibly fuzzy) set of states. A linguistic behavior specifies a probabilistic speaker rule (a
probabilistic choice of message for each state) and a probabilistic hearer rule (a
probabilistic choice of state for each message) given a lexicon. There are many ways of making
these general notions concrete. Since our main emphasis is on how different kinds of language
can be embedded within the replicator mutator dynamic, we focus here on a general formulation
of speaker and listener behavior which is as simple as possible but still in line with both a
game-theoretic tradition
\citep[e.g.,][]{benz:2006,benz+vRooij:2007,jaeger:2007b,franke+jaeger:2014} and a Bayesian
tradition
\citep[e.g.,][]{frank+goodman:2012,goodman+stuhlmueller:2013,GoodmanFrank2016:Pragmatic-Langu}. Here
is what we will assume in the remainder of this paper.

@@

In words, an agent's linguistic behavior -- what message to send when or which message to
interpret how -- is defined by (i) her reasoning level and (ii) her lexicon. In what follows,
this is what we identify an agent's type with, in order to analyze under which conditions
combinations of (i) and (ii) emerge through cultural evolution. Intuitively, literal behavior
(level-$0$ reasoning) results from unreflected language use. Such agents produce and comprehend
solely based on the truth-conditions of their lexicon. In a Gricean spirit, pragmatic behavior
(level-$n+1$ reasoning) results from reasoning about rational language use. Such agents produce
and comprehend by reasoning about how they themselves would interpret or use
expressions. Higher order behavior is therefore, even if in tendency rational, quite simple. It
does not assume agents to know their interlocutor's actual type when interacting (e.g., what
lexicon the interlocutor is using); they simply act based on a conjecture about language use
derived from their own (solipsistic) perspective.

@@

Under the replicator dynamic the proportion of type $i$ in a population will increase or
decrease as a function of its relative fitness $f_i$. In the context of language evolution,
fitness is usually associated with the ability to successfully communicate with other language
users from the same population \citep[e.g.,][]{nowak+krakauer:1999,nowak+etal:2000,
  nowak+etal:2002}. Concretely, the fitness of type $i$ is its average expected communicative
success, or \emph{expected utility} (EU), given the relative frequencies $x_j$ of types $j$ in
the current population:
\begin{align*}
  f_i = \sum_j x_j \text{EU}(t_i,t_j)\,.
\end{align*}
In signaling games, the expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
average success of $i$ when talking or listening to $j$. If, as standardly assumed, agents are speakers half of the
time, this yields:
\begin{align*}
  \text{EU}(t_i,t_j) = \nicefrac{1}{2} \, \text{EU}_S(t_i,t_j) + \nicefrac{1}{2} \, \text{EU}_H(t_i,t_j)\,,
\end{align*}
where $\text{EU}_S(t_i,t_j)$ and $\text{EU}_H(t_i,t_j)$ are the expected utilities for $i$ as a
speaker and as a hearer when communicating with $j$, defined as follows, where $n_i$ and $n_j$
are type $i$'s and type $j$'s pragmatic reasoning types and $L_i$ and $L_j$ are their lexica:
\begin{flalign*}
  & \text{EU}_S(t_i,t_j)  = \sum_s P(s)\sum_m S_{n_i}(m \mid s;L_i) \sum_{s'} H_{n_j}(s' \mid m;L_j)
  \delta(s,s') \,, \\
 & \text{EU}_H(t_i,t_j)  = \text{EU}_S(t_j,t_i)\,.
\end{flalign*}
As usual, we assume that agents are cooperative: $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.

@@

\subsubsection{Learnability}
\label{sec:learnability}
Languages are shaped not only by functionalist forces towards greater communicative success. Another
important factor is the fidelity with which linguistic knowledge is transmitted. Among others,
linguistic production can be prone to errors, states or messages may be perceived incorrectly,
and multiple linguistic hypotheses may be compatible with the data learners are exposed to. These sources
of uncertainty introduce variation in the transmission of linguistic knowledge from one
generation to the next. In particular, learning biases in the iterated transmission process can
influence language evolution substantially.

In biological evolution, where types are expressed genetically, transmission infidelity comes
into the picture through infrequent and mostly random mutation and genetic drift
\citep{kimura:1983}. However, an agent's lexicon and pragmatic reasoning behavior is likely not
inherited genetically. They need to be learned from observation, such as in conditional
imitation (see Appendix~\ref{sec:deriv-repl-mutat}).  Concretely, when agents learn from or
imitate type $j$, they observe the overt linguistic behavior of type $j$ and need to infer the
covert type that most likely produced the observed behavior.

@@

Following \citet{griffiths+kalish:2005,griffiths+kalish:2007} we model language acquisition as
a process of Bayesian inference in which learners combine the likelihood of a type producing
the witnessed learning input with prior inductive biases. In a Bayesian setting these biases
can be codified in a prior $P \in \Delta(T)$, which reflects the amount of data a learner
requires to faithfully acquire the language of the teacher
\citep[cf.][450]{griffiths+kalish:2007}. Put differently, types favored by the prior are easier
to learn because they require less data than less favored types to be inferred, even if they
are both compatible with the learning data at hand. The extent of the prior's influence has
been shown to heavily depend on the learning strategy assumed to underlie the inference
process. On the one hand, early simulation results suggested that weak biases could be
magnified by exposing learners to only small data samples \citep[e.g. in][]{brighton:2002}. On
the other, \citeposs{griffiths+kalish:2007} mathematical characterization showed that iterated
learning converges to the prior in the limit, i.e., that the resulting distribution over types
corresponds to the learners' prior distribution and is not influenced by the amount of input
given to them. This difference in predictions can be traced back to differences in the
selection of hypotheses from the posterior. Griffith \& Kalish's convergence to the prior holds
for learners that sample from the posterior. More deterministic strategies such as the adoption
of the type with the highest posterior probability, so-called {\it maximum a posterior
  estimation} (MAP), increase the influence of both the prior and the data
\citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we use a parameter $l\ge1$ to
modulate between posterior sampling and the MAP strategy. When $l = 1$ learners sample from the
posterior. The learners' propensity to maximize the posterior grows as $l$
increases.\footnote{The way the posterior over types is formed and how learners employ it are
  important design choices. There are other alternatives as well. For instance, instead of
  adopting a single type, learners could keep their uncertainty over types and base their
  production behavior on the posterior over types they form \citep{burkett+griffiths:2010}. A
  straightforward operationalization of this idea would then see production defined as a
  two-step process wherein a speaker first samples from her posterior and then produces
  utterances based on the type she sampled, as done in \citet{kirby+etal:2015}. Beyond adding
  complexity to the model, a disadvantage of this approach is that it could happen that a
  speaker that first uttered a Hungarian utterance speaks Quechua in the next, so to speak. We
  stick to the simpler assumption of a single type being acquired, which we now define.}

@@

@@

In the following we consider a specific application of the model from Section~\ref{sec:model}
which allows us to address the question if or when scalar inferences might (not) lexicalize. We
consider what is perhaps one of the simplest non-trivial setups that speak to this matter and
reflect on their limitations in Section~\ref{sec:discussion}. Section
\ref{sec:setup:-stat-conc} first considers a minimal example intended to isolate effects of
replication and mutation and to gain insights into the sense in which semantic meaning and
pragmatic use can be said to co-evolve. We then turn to a larger and more realistic setup in
Section~\ref{sec:setup:-stat-conc}. Section~\ref{sec:simulation-results} describes simulations
for this case and their results.

\section{Co-evolution of semantics \& pragmatics in a restricted type space}
\label{subsec:example}

This section illustrates how replication and mutation act on the evolution of semantic meaning
and pragmatic use in the simplest non-trivial type space. Our goal is to sharpen intuitions
about the sense in which semantics and pragmatics might be said to co-evolve.

  % \path[->, thick, densely dashdotted, Steelblue] (0.75,0.75) edge [bend left] node[sloped,below] {\footnotesize learning\textsubscript{flat}} (2.25,2.45);
  % \path[->, thick, densely dashdotted, Steelblue] (4.25,0.75) edge [bend right] node[sloped,below] {\footnotesize learning\textsubscript{flat}} (2.75,2.45);
  % \path[->, thick, densely dashdotted, Steelblue] (0.75,4.25) edge [bend right] node[sloped,above] {\footnotesize learning\textsubscript{flat}} (2.25,2.75);
  % \path[->, thick, densely dashdotted, Steelblue] (4.25,4.25) edge [bend left] node[sloped,above] {\footnotesize learning\textsubscript{flat}} (2.75,2.75);

  %   \draw[->, thick, goldish] (0.75,-0.25) -- (2.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
  %   \draw[<-, thick, goldish] (2.75,-0.25) -- (4.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
  %   \draw[->, thick, goldish] (-0.25,0.75) -- (-0.25,4.25) node[midway, above,rotate=90] {\footnotesize learning\textsubscript{biased}};
  %   \draw[->, thick, goldish] (5.25,0.75) -- (5.25,4.25) node[midway, below,rotate=90] {\footnotesize learning\textsubscript{biased}};
  %   \draw[->, thick, goldish] (0.75,5.25) -- (4.25,5.25) node[midway, above] {\footnotesize learning\textsubscript{biased}};

  %   \end{tikzpicture}
\caption{Sketch of dynamics on a two-dimensional type space with four types. The $y$-axis
  represents the proportion of types with $\Llack$; the $x$-axis that of pragmatic ones. Arrows
  sketch population trajectories under functional pressure (replicator steps, red arrows) and
  learnability pressure (mutator steps) with either a flat prior (blue dash-dotted arrows) or a
  prior that favors $\Llack$ (solid purple arrows).}
\label{fig:sketch}
\end{figure}

We focus only on the four types from Section~\ref{sec:languages+use}. Recall that these types
are made up of a combination of either lexicon $\Llack$ or $\Lbound$ and either literal
(level-$0$) or pragmatic (level-$1$) behavior.

The population under consideration consists of only the four types from
Section~\ref{sec:languages+use}. Recall that these types are made up of a combination of either
lexicon $\Llack$ or $\Lbound$ and either literal (level-$0$) or pragmatic (level-$1$)
behavior. We may think of such a population as a point in a two-dimensional space. The grey
dotted square in Fig.~\ref{fig:sketch} outlines this space. The space's first dimension
determines the proportion of types that lexicalize ambiguous $\msome$ (the $y$-axis in
Fig.~\ref{fig:sketch}). That is, the proportion of types with $\Llack$. The second dimension
determines the proportion of pragmatic language users (the $x$-axis in
Fig.~\ref{fig:sketch}). Populations at the corners of this space are monomorphic; all agents in
them are of a single type. Points away from the corners represent mixed populations of various
types.\footnote{This two-dimensional representation underspecifies the exact proportions of
  types found in a mixed population. The space's center, for instance, could be composed 50\%
  of pragmatic $\Llack$ and of an equal proportion of literal $\Lbound$. But it could equally
  well be a population that has each type with the same frequency. This level of granularity
  will nevertheless be sufficient for our purposes, as we are only interested in sketching
  rough trajectories after replicator or mutator steps.}

From Section~\ref{sec:languages+use}, we already know some things about this type space. When
it comes to communicative success: types with $\Lbound$ have a functional advantage over
$\Llack$; literal $\Lbound$ has a slight advantage over pragmatic $\Lbound$; all have a sizable
advantage over literal $\Llack$. The red arrows in Fig.~\ref{fig:sketch} sketch out these
relationships. They depict the rough trajectories we can expect if there is only pressure for
communicative success. How much each replicator step moves a population through this space
ultimately depends on the rationality parameter $\lambda$. In particular, note that functional
differences decrease as $\lambda$ increases: if $\lambda$ is sufficiently high, all types
except for literal $\Llack$ speakers will show a strong tendency to associate $\ssome$ with
$\msome$ and $\sall$ with $\mall$.

As a consequence, when it comes to learnability and transmission fidelity, the likelihood that
any of these three types produces a given datum $d$ will be approximately equal. We can
therefore expect that, if the prior is flat, no type is passed on very faithfully in this
space. This would lead populations to gravitate toward the space's center after mutator
steps. This is sketched by the blue dash-dotted arrows in Fig.~\ref{fig:sketch}. Neither prior
nor likelihood alone, but their combination, are what determines the effects of (iterated)
Bayesian learning. If, instead of a flat prior, there is a prior that favors, say, $\Llack$
over $\Lbound$ we can expect mutator steps to be skewed toward the upper-half of the
space. This is suggested by the solid purple arrows in Fig.~\ref{fig:sketch}. Taking stock, the
effects of iterated learning will depend not only on $\lambda$, affecting the likelihood, but
also on the prior (how strong of a bias there is) and the degree to which learners tend to
maximize the posterior (controlled by parameter $l$).

\begin{figure}[t]
\centering
\includegraphics[width=1.08\textwidth,height=10cm, keepaspectratio]{./plots/fig-contourb105k5.png}
\caption{Dynamics on a two-dimensional type space with four types. The $y$-axis represents the
  proportion of types with $\Llack$; the $x$-axis that of pragmatic ones. Arrows indicate
  directionality of trajectories after (a) replicator steps and (b) mutator steps with a prior
  that favors types with $\Llack$ by a factor of $1.05$ over those with $\Lbound$. Colored
  contours show the proportion of pragmatic $\Llack$ users after a step. \tb{I don't think the
    contours are adding much, so they could be removed. Then again, I think they are rather
    non-intrusive, look nice, and illustrate how the step-size, depending on where we are in
    the space, varies.} \mf{I think that the contours are sweet. But I worried about the size
  of the arrows. I made the picture a bit bigger}}

Fig.~\ref{fig:quiver} shows actual evolutionary trajectories in this type space. The replicator
steps make the functional disadvantage of $\Llack$ against $\Lbound$ apparent, particularly for
its literal users and/or low $\lambda$. They also highlight that the functional difference
between literal and pragmatic $\Lbound$ is small. Under functional pressure only, populations
move toward the lower part of the square but the steps become small as the lower edge,
representing populations composed only of $\Llack$ users, are reached. As for the mutator
diagrams, there are two main things to note. As expected, a small bias for a type will lead to
larger proportions to adopt it, provided that likelihood offers little to no differentiation
between types (see lower left space in Figure~\ref{fig:quiver}). However, the bias is far from
absolute. If, instead, certain types are more likely to be recovered based on the linguistic
input they produce, they will be passed on more faithfully. This is why, for instance,
populations move toward a majority of $\Llack$-types for $\lambda = 5$ and $l = 10$.\mf{last
  sentence: I don't get it; looks to me as if they are going the other direction; only in the
  lower right case do I see a shift in that direction}

Fig.~\ref{fig:sketch} and \ref{fig:quiver} depict replicator and mutator steps in isolation to
better showcase their effects. Their combination will yield different outcomes, which we now
turn to in a case study with a larger type space and more motivation for particular assumptions
about learning biases.
